Wait-free Queues with Polylogarithmic Step Complexity

### 2022-10-17 thesis/thesis.tex
finalize description notes

In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently. We design a queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of invocation. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue. Unlike their work, our algorithm does not use \texttt{CAS} operations with big words and does not put a limit on the number of concurrent dequeuers.

@@

To propagate operations to node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh()} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps other processes doing \nf{Refresh} on $n$ concurrently to propagate their operations up to the parent. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
This is because if both of the \tt{Refresh}es on $n$ fail to append then there is another instance of \nf{Refresh} in between which has gathered its new operations from $n$'s children after the first failed \nf{Refresh} and succeeded to do an append on $n$. This \nf{Refresh} helps to append the new operations in the $n$' children before the first failed \nf{Refresh}, in case they were not already appended.

@@

The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh}. This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$' second failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}). We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph. After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to root two times on each node. So with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization. This cooperative solution allows us to overcome the CAS Retry Problem.

It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. We can only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in root (see Figure \ref{fig::set}). Achieving a constant sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations in a \nf{Refresh} step. It contains the number of operations from the left and the right child propagated to the node by a \texttt{Refresh} procedure (see Figure \ref{fig:block} for an example). A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parents' blocks. We call the aggregated blocks, subblocks of the new block and the new block, superblock of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering we can design a universal construction; for given object $O$ and operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then return the response for $op$. However, this approach is not enough for an efficient queue.
A process may wish to know (1) the $i$th propagated operation or (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) and use them to construct a fast queue below.

After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblock of $B$ in a child of the root that contains that $i$th operation. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.
To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in. After a block is installed in a node we store the approximate index of its superblock in it to make this faster.

@@

In an execution on a queue where no dequeue operation returns\nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns\nf{null} if and only if the size of the queue after the previous operation is 0. We refer to such a dequeue as a\nf{null} dequeue. Otherwise, if the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues up to block B. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue which is the response to a given dequeue in constant time.

In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific rank.  We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. Consider the follwing operations, where operations in a cell are concurrent.
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\end{table}

@@

Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$. Otherwise the response would be \nf{null}.

\subsection{Details of the Implementation}
Pseudocode for the queue implementation is given in Section \ref{algQ}.
\paragraph{Node} 
 In each \texttt{Node} we store pointers to its parent and children, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{Block}
 The information stores in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf. If it is in a leaf, we use a \nf{LeafBlock} which simply stores one operation. If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$. The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the block prior to $B$ ended. In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. \texttt{sum\textsubscript{enq-left}} is the total number of \nf{Enqueue} operations in theblocks before the last subblock of $B$ in the left child child, other fields semantics is also similar. The \nf{end\sub{left}} and \nf{end\sub{right}} field store the last subblock of a block in the left and the right child, respectively. The approximate index of the superblock of non-root blocks is stored in their \nf{super} field. The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.

@@

\paragraph{\tt{Dequeue()}}
\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields, appends it to the tree. Then it computes the position of the appended \nf{Dequeue} operation in the root and after that finds the response of the \nf{Dequeue} calling \nf{FindResponse}.

\paragraph{\tt{FindResponse($b, i$)}}
To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not. If there are more \nf{Dqeueue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dqeueue}. If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is response to the \nf{Dqeueue}. Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block we find the block) and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.

\paragraph{\tt{Append($B$)}}
The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}. \nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}. When \nf{Propagate} terminates it is guaranteed that the appended block is a subblock of a block in the \nf{root}. There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it.

\paragraph{\tt{Propagate()}}
\nf{Propagate} on node $n$ uses the double refresh idea described in Section \ref{DescriptQ} and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively untill it reaches the root.

\paragraph{\tt{Refresh()}}
The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}. The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recent appended blocks. The reason behind this helping is explained later when we discuss \nf{IndexDequeue}. After helping to \nf{Advance} children, \nf{new} block is created in Line \ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there is no new operations to propagate and it is unnecessary to add an empty block to the tree. Later we will use the fact that all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}. If it was successful all is good. If not, it means someone else has already put a block in $n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps update $n$\nf{.head} and the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
The \nf{new} block is created in Line 333 and its fields are filled similarly for both left and right directions. The variable  \nf{index\sub{prev}} is the index of the block preceding the first direct subblock in direction \nf{dir} aggregated into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in \nf{dir} child. Then \nf{sum\sub{enq-dir}} is computed from sum of the the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new} block is going to be installed in the \nf{root}, the \nf{size} field is also computed.

\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DSearch($e, end$)}}
We can show an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation in that block. If we know the block and rank within the block of an operation we can find the subblock containig the operation and the operation's rank within that subblock in poly-log time. To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}. \nf{DSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
A call to \nf{$n$.GetEnqueue(b,i)} returns the \nf{element} of the \nf{i}th enqueue in \nf{b}th the block of $n$. The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block. Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).

@@

$n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block of the $i$th \nf{Dequeue} of $b$th block of $n$ in the root. Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}. Let $n$\nf{.parent} be $n$\nf{.parent}. Without loss of generality assume for the rest of this section a ${n}$ is the left child of $n$\nf{.parent}. Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${b}$ for \nf{left.head} therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}. Let the index of the block $R_p$ put in $n$\nf{.parent}\nf{.blocks} be $j$.

Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating it. One approach for $R_n$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $n$\nf{.parent}. This solution would not be efficient because there might be $n$\nf{.parent} subblocks in the block $R_p$ propagated needing updates their \nf{super} field. However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation. If we know the approximate position of the superblock of $B$ then we can search for the real superblock when we it is needed. Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$. We can set \nf{$B$.super} to \nf{$n$.parent.head} while creating $B$, but the problem is that there might be many \nf{$n$\nf{.parent}.Refresh}es that could happen after $R_n$ reads $n$\nf{.parent}\nf{.head} and before propagating $B$ to $n$\nf{.parent}. If $R_n$ sets \nf{$B$.super} to \nf{$n$\nf{.parent}.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}. In this case, the next \nf{Refresh}es on $n$ and \nf{n.parent} help fill in the value of \nf{$B$.super}.

Block $B$ is appended to \nf{$n$.blocks[i]} on Line \ref{cas}. After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}. We bound how far \nf{b.super} is from the index of $b$'s superblock by 1.

>>>>

In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently. We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.

Jayanti and Petrovic introduced a wait-free poly-logarithmic multi-enqueuer single-dequeuer queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having a tournament tree among processes to agree on the linearization of operations to design a a polylogarithmic multi-enqueuer multi-dequeuer queue. Unlike their work, our algorithm does not put a limit on the number of concurrent dequeuers.

@@

To propagate operations to a node $n$ in the tree, a process observes the operations in both of $n$'s children that are not already in $n$, merges them to create an ordering and then tries to append the ordering to the sequence stored in $n$. We call this procedure \tt{$n$.Refresh()} (see Figure \ref{fig::propagstep}). A \nf{Refresh} on $n$ with a successful append helps to propagate their operations up to the $n$. We shall prove that if a process invokes \tt{Refresh} on the node $n$ two times and fails to append the new operations to $n$ both times, the operations that were in $n$'s children before the first \tt{Refresh} are guaranteed to be in $n$ after the second failed \tt{Refresh}.
We sketch the argument here.

@@

We use \nf{CAS} (Compare \& Swap) instructions to implement the \nf{Refresh}'s attempt to append  described in the previous paragraph.
The second failed \nf{Refresh} of $P$ is assuredly concurrent with a successful \texttt{Refresh} that has read its information after the invocation of the first failed \nf{Refresh} (see Figure \ref{fig::simpleDoubleRefresh}). This is because some process $L$ does a successful append during $P$'s first failed attempt, and some process $K$ performs a \nf{Refresh} that reads its information after $L$'s append and then performs a successful append during $P$'s second failed \nf{Refresh}. Process $K$'s \nf{Refresh} helps to append the new operations in $n$'s children before $P$'s first failed \nf{Refresh}, in case they were not already appended. After a process appends its operation into its leaf it can call \nf{Refresh} on the path up to the root at most two times on each node. So, with $O(\log p)$ \nf{CAS}es an operation can ensure it appears in the linearization. This cooperative solution allows us to overcome the CAS Retry Problem.

It is not efficient to store the sequence of operations in each node explicitly because each operation would have to be copied all the way up to the root; doing this would not be possible in poly-logarithmic time.
Instead we use an implicit representation of the operations propagated together. Furthermore, we do not need to maintain an ordering on operations propagated together in a node until they have reached the root. It is sufficient to only keep track of sets of operations propagated together in each \nf{Refresh} and then define the linearization ordering only in the root (see Figure \ref{fig::set}). Achieving a constant-sized implicit representation of operations in a \nf{Refresh} allows us to \texttt{CAS} fixed-size objects in each \tt{Refresh}. To do that, we introduce \it{block}s. A block stores information about the operations propagated by a \nf{Refresh}. It contains the number of operations from the left and the right child propagated to the node by the \texttt{Refresh} procedure. See Figure \ref{fig:block} for an example. A node stores an array of blocks of operations propagated up to it. A propagate step  aggregates the new blocks in children into a new block and puts it in the parent's blocks. We call the aggregated blocks \it{subblocks} of the new block and the new block the \it{superblock} of them.  In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Thus, there are at most $p$ operations in a block. Furthermore, since the operations in a \texttt{Refresh} step are concurrent we can linearize them  among themselves in any order we wish, because if two operations are read in one successful \tt{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

So far, we have a shared tree that processes use to agree on the implicit ordering stored in its root. With this agreement on linearization ordering we can design a universal construction; for given object $O$ we can perform an operation $op$ by applying all the operations up until $op$ in the root on a local copy of the object and then returning the response for $op$. However, this approach is not enough for an efficient queue.
We show that we can build an efficien queue if we can compute two things about the ordering in the root: (1) the $i$th propagated operation and (2) the rank of a propagated operation in the linearization. We explain how to implement (1) and (2) in poly-logarithmic steps.

After propagating an operation \texttt{op} to the root, processes can find out information about the linearization ordering using (1) and (2). 
To get the $i$th operation in the root, we find the block $B$ containing the $i$th operation in the root, and then recursively find the subblocks of $B$ in the descendent of the root that contain that $i$th operation. When we reach a block in a leaf, the operation is explicitly stored there. To make this search faster, instead of iterating over all blocks in the node, we store the prefix sum of the number of elements in the blocks sequence to permit a binary search for the required block. We also store pointers to determine the range of subblocks of a block to make the binary search faster. In each block, we store the prefix sum of operations from the left child and from the right child. Moreover, for each block, we store two pointers to the last left and right subblock of it (see Figure \ref{fig::pointer}). We know a block size is at most $p$, so binary search takes at most \textsc{O}$(\log p)$ time, since the pointers of a block and its previous block reduce the search range size to \textsc{O}$(p)$.

To compute the rank in the root of an operation in the leaf, we need to find the superblock of the block that operation is in. After a block is installed in a node we store the approximate index of its superblock in it to make this faster.

@@

In an execution on a queue where no dequeue operation returns\nf{null}, the $k$th dequeue returns the argument of the $k$th enqueue. In the general case a dequeue returns\nf{null} if and only if the size of the queue after the previous operation is 0. We refer to such a dequeue as a \it{null dequeue}. If the the dequeue is the $k$th non-null dequeue, it returns the argument of the $k$th enqueue. Having the size of the queue after an operation we can compute the number of non-null dequeues from the number of enqueues up to block B. So, if we store the size of the queue after each block of operations in the root, we can compute the index of the enqueue whose argument is the response to a given dequeue in constant time.

In our case of implementing a queue, a process only needs to compute the rank of a \nf{Dequeue} and get an \nf{Enqueue} with a specific rank.  We know we can linearize operations in a block in any order; here, we choose to put \nf{Enqueue} operations in a block before \nf{Dequeue} operations. Consider the following operations, where operations in a cell are concurrent.
\begin{table}[hbtp]
\centering
\begin{small}
\begin{tabular}{c|c|c|c}
    \hline \texttt{Deq} & \texttt{Enq(5)}, \texttt{Enq(2)}, \texttt{Enq(1)}, \texttt{Deq}& \texttt{Enq(3)}, \texttt{Deq}&  \texttt{Enq(4)}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}, \texttt{Deq}\\ \hline
  \end{tabular}
\end{small}
\end{table}

@@

Given a \texttt{Dequeue} is in block $B$, its response is the argument of the \tt{Enqueue} whose rank is the number of non-null \nf{Dequeue}s in  blocks $1,2,...,b-1 + $ index of the \nf{Dequeue} in $B$\textrm{'s} \nf{Dequeue}s, if $\big ($size of the queue after $b-1\textrm{th block} + \#\textrm{\tt{Enqueue}s in }b\textrm{th block} - \#\textrm{index of \tt{Dequeue} in }B\textrm{'s \tt{Dequeue}s} \big )\geq 0$. Otherwise the response would be \nf{null}.

\subsection{Details of the Implementation}
Pseudocode for the queue implementation is given in Section \ref{algQ}. It uses the following two types of objects.
\paragraph{\tt{Node}} 
 In each \texttt{Node} we store pointers to its parent and children, an array of \nf{Block}s called \texttt{block}s and the index \nf{head} of the first empty entry in \texttt{blocks}.

\paragraph{\tt{Block}}
 The information stored in a \nf{Block} depends on whether  the \nf{Block} is in an internal node or a leaf. If it is in a leaf, we use a \nf{LeafBlock} which simply stores one operation. If a block $B$ is in an internal node $n$, then it contains subblocks in the left and right children of $n$. The left subblocks of $B$ are some consecutive blocks in the left child of $n$ starting from where the block prior to $B$ ended. In each \texttt{block} we store four essential fields that implicitly summarize which operations are in the block \texttt{sum\textsubscript{enq-left}}, \texttt{sum\textsubscript{deq-left}}, \texttt{sum\textsubscript{enq-right}}, \texttt{sum\textsubscript{deq-right}}. The \texttt{sum\textsubscript{enq-left}} field is the total number of \nf{Enqueue} operations in the blocks before the last subblock of $B$ in the left child. The other fields' semantics are similar. The \nf{end\sub{left}} and \nf{end\sub{right}} field store the last subblock of a block in the left and the right child, respectively. The approximate index of the superblock of non-root blocks is stored in their \nf{super} field. The \texttt{size} field in a block in the root node stores the size of the queue after the operations in the block have been performed.

@@

\paragraph{\tt{Dequeue()}}
\nf{Dequeue} creates a \nf{LeafBlock}, sets its \nf{sum\sub{enq}} and \nf{sum\sub{deq}} fields and appends it to the tree. Then, it computes the position of the appended \nf{Dequeue} operation in the root using \nf{IndexDequeue} and after that finds the response of the \nf{Dequeue} by calling \nf{FindResponse}.

\paragraph{\tt{FindResponse($b, i$)}}
To compute the response of the $i$th \nf{Dequeue} in the $b$th block of the root Line \ref{checkEmpty} computes whether the queue is empty or not. If there are more \nf{Dequeue}s than \nf{Enqueue}s the queue would become empty before the requested \nf{Dequeue}. If the queue is not empty, Line \ref{computeE} computes the rank $e$ of the \nf{Enqueue} whose argument is response to the \nf{Dequeue}. Knowing the response is the $e$th \nf{Enqueue} in the root (which is before the $b$th block) we find the block and position containing the \nf{Enqueue} operation using \nf{DSearch} and after that \nf{GetEnqueue} finds its \nf{element}.

\paragraph{\tt{Append($B$)}}
The \nf{head} field is the index of the first empty slot in \nf{blocks} in a \nf{LeafBlock}. There are no multiple write accesses on \nf{head} and \nf{blocks} in a leaf because only the process that the leaf belongs to appends to it. \nf{Append($B$)} adds $B$ to the end of the \nf{blocks} field in the leaf, increments \nf{head} and then calls \nf{Propagate} on the leaf's \nf{parent}. When \nf{Propagate} terminates it is guaranteed that the appended block is a subblock of a block in the \nf{root}.

\paragraph{\tt{Propagate()}}
\nf{Propagate} on node $n$ uses the double refresh idea described in Section \ref{DescriptQ} and invokes two \nf{Refresh}es on $n$ in Lines \ref{firstRefresh} and \ref{secondRefresh}. Then, it invokes \nf{Propagate} on \nf{$n$.parent} recursively until it reaches the root.

\paragraph{\tt{Refresh()}}
The goal of a \nf{Refresh} on node $n$ is to create a block of $n$'s children's new blocks and append it to $n$\nf{.blocks}. The variable \nf{h} is read from $n$\nf{.head} at Line \ref{readHead}. The new block created by \nf{Refresh} will be inserted into $n\nf{.blocks[h]}$. Lines \ref{startHelpChild1}--\ref{endHelpChild1} of \nf{$n$.Refresh} help to \nf{Advance} $n$'s children. \nf{Advance} increments the children's \nf{head} if necessary and sets the \nf{super} field of their most recent appended blocks. The reason behind this helping is explained later when we discuss \nf{IndexDequeue}. After helping to \nf{Advance} the children, a new block called \nf{new} is created in Line \ref{invokeCreateBlock}. Then, if \nf{new} is empty, \nf{Refresh} returns \nf{true} because there are no new operations to propagate and it is unnecessary to add an empty block to the tree. Later we will use the fact that all blocks contain at least one operation. Line \ref{cas} tries to install \nf{new}. If it was successful all is good. If not, it means someone else has already put a block in $n$\nf{.blocks[h]}. In this case, \nf{Refresh} helps advance $n$\nf{.head} to \nf{h+1} and update the \nf{super} field of $n$\nf{.blocks[h]} at Line \ref{advance}.

\paragraph{\tt{CreateBlock()}} \texttt{$n$.CreateBlock($h$)} is used by \nf{Refresh} to construct a block containing new operations of $n$'s children.
The block \nf{new} is created in Line 333 and its fields are filled similarly for both left and right directions. The variable  \nf{index\sub{prev}} is the index of the block preceding the first subblock in the child in direction \nf{dir} that is aggregated into \nf{new}. Field \texttt{new.end\textsubscript{dir}} stores the index of the rightmost subblock of \nf{new} in the child. Then \nf{sum\sub{enq-dir}} is computed from sum of the the number of \nf{Enqueue} operations in the \nf{new} block from direction \nf{dir} and the value stored in \texttt{$n$.blocks[h-1].sum\sub{enq-dir}}. The field \nf{sum\sub{deq-dir}} is computed similarly. Then, if \nf{new} is going to be installed in the \nf{root}, the \nf{size} field is also computed.

\paragraph{\tt{GetEnqueue($b,i$)} and \tt{DSearch($e, end$)}}
We can describe an operation in a node in two ways: the rank of the operation among all the operations in the node or the index of the block containing the operation in the node and the rank of the operation within that block. If we know the block and rank within the block of an operation we can find the subblock containing the operation and the operation's rank within that subblock in poly-log time. To find the response of a \nf{Dequeue}, we know about the  rank of the response \nf{Enqueue} in the root (\nf{e} in Line \ref{computeE}).
We also know the \nf{e}th \nf{Enqueue} is in \nf{root.blocks[1..end]}. \nf{DSearch} uses doubling to find the range that contains the answer block (Lines \ref{dsearchStart}--\ref{dsearchEnd}) and then tries to find the required indices with a binary search (Line \ref{dsearchBinarySearch}).
A call to \nf{$n$.GetEnqueue($b,i$)} returns the \nf{element} of the $i$th enqueue in $b$th the block of $n$. The range of subblocks of a block is determined using the \nf{end\sub{left}} and \nf{end\sub{right}} fields of the block and its previous block. Then, the subblock is found using binary search on the \nf{sum\sub{enq}} field (Lines \ref{leftChildGet} and \ref{rightChildGet}).

@@

A call to $n$\nf{.IndexDequeue($b,i$)} computes the block and the rank within the block in the root of the $i$th \nf{Dequeue} of the $b$th block of $n$. Let $R_n$ be the successful  \nf{Refresh} on node $n$ that did a successful \nf{CAS(null, $B$)} into \nf{n.blocks[$b$]}. Let $par$ be $n$\nf{.parent}. Without loss of generality assume for the rest of this section a ${n}$ is the left child of $par$. Let $R_{par}$ be the first successful \nf{$par$.Refresh} that reads some value greater than ${b}$ for \nf{left.head} and therefore contains ${B}$ in its created block in Line \ref{invokeCreateBlock}. Let $j$ be the index of the block that $R_{par}$ put in $par$\nf{.blocks}.

Since the index of the superblock of ${B}$ is not known until $B$ is propagated, $R_n$ cannot set the \nf{super} field of ${B}$ while creating it. One approach for $R_{par}$ is to set the \nf{super} field of ${B}$ after propagating $B$ to $par$. This solution would not be efficient because there might be up to $p$ subblocks in the block $R_{par}$ propagated needing updates their \nf{super} field. However, intuitively, once $B$ is installed, its superblock is going to be close to \nf{$n$.parent.head} at the time of installation. If we know the approximate position of the superblock of $B$ then we can search for the real superblock when it is needed. Thus, \nf{$B$.super} does not have to be the exact location of  the superblock of $B$, but we want it to be close to $j$. We can set \nf{$B$.super} to \nf{$par$.head} while creating $B$, but the problem is that there might be many \nf{Refresh}es on $par$ that could happen after $R_n$ reads $par$\nf{.head} and before propagating $B$ to $par$. If $R_n$ sets \nf{$B$.super} to \nf{$par$.head} after appending $B$ to \nf{$n$.blocks} (Line \ref{setSuper1}), $R_n$ might go to sleep at some time after installing $B$ and before setting \nf{$B$.super}. In this case, the next \nf{Refresh}es on $n$ and $par$ help fill in the value of \nf{$B$.super}.

Block $B$ is appended to \nf{$n$.blocks[$b$]} on Line \ref{cas}. After appending $B$, \nf{$B$.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or by Line \ref{helpAdvance} of a \nf{$n$.parent.Refresh}. We shall show that this is sufficient to ensure that \nf{$B$.super} differs from the index of $B$'s superblock by at most 1.
